---
title: "Machine Learning Project"
author: "treepruner"
date: "October 9, 2015"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Project Description
Sensors were placed on the belt, forearm, arm and dumbell of 6 participants. These sensors detected the acceleration, gyroscope and magnetometer data at a joint sampling rate of 45 Hz. The participants were then asked to lift a barbell 5 different ways: 
 * Class A exactly according to the specification 
 * Class B throwing the elbows to the front
 * Class C lifting the dumbbell only halfway 
 * Class D lowering the dumbbell only halfway 
 * Class E throwing the hips to the front 
 
Our task is to develop a model with only the accelerometer readings to identify which class method was used to lift
the dumbell. 

The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har. 

### Executive Summary
The rpart method produced a model with an accuracy of XXX. The most useful variables proved to be  

#### Load Packages 
```{r, echo = FALSE, message=FALSE}
library(dplyr)
library(grid)
library(gridExtra)
library(caret)
library(rattle)
library(party)
library(partykit)
library(doParallel)
```

#### Download Data
```{r}
#if (!file.exists("project")) { dir.create("project")} 
#setwd("./project")

#setInternet2(use = TRUE)
#trainURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
#download.file(trainURL, "pml_training.csv", method = "auto")

#testURL <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
#download.file(testURL, "pml_testing.csv", method = "auto")
```

#### Load Training Data Set and Check Control Totals
Control Totals per Forum:
|    A|    B|    C|    D|    E| 
|=====||=====||=====||=====||=====|
| 5580| 3797| 3422| 3216| 3607| 

My Training Totals:
```{r }
pml_training <- read.csv("pml_training.csv", na.strings = c("#DIV/0!", "NA"))
table(pml_training$classe)
```



##### Subset for classe and _X, _Y, _Z Variables Only

Exclude username, timestamps, windows and classe variables
Exclude all the variables calculated from the original sensor measurements identified with variable names containing accel, gyros, magnet and _x, _y or _z.


```{r }
xyz <- select(pml_training, classe, contains("_x"), contains("_y"), contains( "_z"), -contains("stddev"), -contains("kurtosis"), -contains("max"), -contains("min"), -contains("yaw") )

```

#### Explore _xyz Data with Correlations

```{r}
corXYZ <- abs(cor(xyz[, -1]))
diag(corXYZ) <- 0
corXYZ <- which(corXYZ > 0.8, arr.ind = T)
corXYZ[order(rownames(corXYZ)), ] 
```

#### Combine Highly Correlated Variables

```{r}
xyz$accel_belt_xy <- xyz$accel_belt_x + xyz$accel_belt_y
xyz$gyros_arm_xy <- xyz$gyros_arm_x  + xyz$gyros_arm_y 
xyz$gyros_dumbbell_xz <- xyz$gyros_dumbbell_x + xyz$gyros_dumbbell_z
xyz$gyros_forearm_yz <- xyz$gyros_forearm_y + xyz$gyros_forearm_z
xyz$magnet_arm_xyz <- xyz$magnet_arm_x + xyz$magnet_arm_y + xyz$magnet_arm_z
xyz$accelMagnet_belt_x <- xyz$accel_belt_x + xyz$magnet_belt_x
```

#### Remove Highly Correlated Variables

```{r}
xyz <- select(xyz, 
              -accel_belt_x, -accel_belt_y,
              -gyros_arm_x, -gyros_arm_y,
              -gyros_dumbbell_x, -gyros_dumbbell_z,
              -gyros_forearm_y, -gyros_forearm_z,
              -magnet_arm_x, -magnet_arm_y, -magnet_arm_z,
              -accel_belt_x,-magnet_belt_x)
```


### Create _xyz Training and Validation Set

I used the recommended value of .6 to create a training set which leaves 40% for a validation set in order to calculate an out of sample error rate.
```{r }
set.seed(1138)
inTrain_xyz <- createDataPartition(y = xyz$classe, p = 0.6, list = FALSE)
training_xyz <- xyz[inTrain_xyz,]
validation_xyz <- xyz[-inTrain_xyz,]
```


\pagebreak
### Explore _xyz Acceleration Variables Graphically


```{r, fig.width = 6, fig.height=2, echo = FALSE }
xAccelMagnetBelt <- qplot(accelMagnet_belt_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xAccelArm <- qplot(accel_arm_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xAccelForearm <- qplot(accel_forearm_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xAccelDumbbell <- qplot(accel_dumbbell_x, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(xAccelMagnetBelt, xAccelArm, xAccelForearm, xAccelDumbbell, ncol = 4)

xyAccelBelt <- qplot(accel_belt_xy, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
yAccelArm <- qplot(accel_arm_y, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
yAccelForearm <- qplot(accel_forearm_y, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
yAccelDumbbell <- qplot(accel_dumbbell_y, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(xyAccelBelt, yAccelArm, yAccelForearm, yAccelDumbbell, ncol = 4)

zAccelBelt <- qplot(accel_belt_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zAccelArm <- qplot(accel_arm_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zAccelForearm <- qplot(accel_forearm_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zAccelDumbbell <- qplot(accel_dumbbell_z, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(zAccelBelt, zAccelArm, zAccelForearm, zAccelDumbbell, ncol = 4)
```
\pagebreak

### Explore _xyz Gyroscope Variables Graphically

```{r, fig.width = 6, fig.height=2, echo = FALSE  }
xgyrosBelt <- qplot(gyros_belt_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xygyrosArm <- qplot(gyros_arm_xy, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xgyrosForearm <- qplot(gyros_forearm_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xzgyrosDumbbell <- qplot(gyros_dumbbell_xz, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(xgyrosBelt, xygyrosArm, xgyrosForearm, xzgyrosDumbbell, ncol = 4)

ygyrosBelt <- qplot(gyros_belt_y, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xygyrosArm <- qplot(gyros_arm_xy, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
yzgyrosForearm <- qplot(gyros_forearm_yz, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
ygyrosDumbbell <- qplot(gyros_dumbbell_y, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(ygyrosBelt, xygyrosArm, yzgyrosForearm, ygyrosDumbbell, ncol = 4)

zgyrosBelt <- qplot(gyros_belt_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zgyrosArm <- qplot(gyros_arm_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
yzgyrosForearm <- qplot(gyros_forearm_yz, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xzgyrosDumbbell <- qplot(gyros_dumbbell_xz, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(zgyrosBelt, zgyrosArm, yzgyrosForearm, xzgyrosDumbbell, ncol = 4)
```

\pagebreak
### Explore _xyz Magnetometer Variables Graphically

```{r, fig.width = 6, fig.height=2, echo = FALSE  }
xaccelMagnetBelt <- qplot(accelMagnet_belt_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xyzmagnetArm <- qplot(magnet_arm_xyz, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xmagnetForearm <- qplot(magnet_forearm_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xmagnetDumbbell <- qplot(magnet_dumbbell_x, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(xaccelMagnetBelt, xyzmagnetArm, xmagnetForearm, xmagnetDumbbell, ncol = 4)

ymagnetBelt <- qplot(magnet_belt_y, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xyzmagnetArm <- qplot(magnet_arm_xyz, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xmagnetForearm <- qplot(magnet_forearm_x, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
ymagnetDumbbell <- qplot(magnet_dumbbell_y, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(ymagnetBelt, xyzmagnetArm, xmagnetForearm, ymagnetDumbbell, ncol = 4)

zmagnetBelt <- qplot(magnet_belt_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
xyzmagnetArm <- qplot(magnet_arm_xyz, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zmagnetForearm <- qplot(magnet_forearm_z, colour = classe, classe, data = training_xyz) + guides(colour = FALSE)
zmagnetDumbbell <- qplot(magnet_dumbbell_z, colour = classe, classe,  data = training_xyz) + guides(colour = FALSE)
grid.arrange(zmagnetBelt, xyzmagnetArm, zmagnetForearm, zmagnetDumbbell, ncol = 4)
```



### Run rpart xyz 

```{r}
trCtrl <- trainControl(method="repeatedcv",             
                     repeats=5,                          
                     classProbs=TRUE) # accuracy : 0.4842

set.seed(1138)
registerDoParallel(4)       
getDoParWorkers()

xyzFit <- train(classe ~., data = training_xyz,
             method = "rpart",
             control = rpart.control(minsplit = 2),
             trControl = trCtrl,
             tuneLength = 5)
varImp(xyzFit, compete= FALSE)
```


```{r, fig.width = 3, fig.height=3 }
dotPlot(varImp(xyzFit, compete= FALSE))
```

```{r, fig.width = 8, fig.height=6 }
fancyRpartPlot(xyzFit$finalModel, sub="")
```


#### Compare with Valadation xyz
```{r}
xyzFitPred <- predict(xyzFit, newdata = validation_xyz) 
confusionMatrix(xyzFitPred, validation_xyz$classe)
```


### Load Testing Data Set
```{r }
pml_testing <- read.csv("pml_testing.csv", 
                        na.strings = c("#DIV/0!", "NA"))
dim(pml_testing)

```

